Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-03 paper/tidy-missing-data-paper.Rmd
fixes section/table/figure referencing with bookdown::pdf_book: base_format: rticles::jss_article

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
  # For running headers, if needed
  short:     "\\pkg{naniar}: Explore missing data"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [kworkflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output: 
  rticles::jss_article:
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

housing_raw <-
  readr::read_csv("../data/melbourne_housing_raw.csv") %>%
  janitor::clean_names() %>%
  dplyr::rename(region_name = regionname,
         property_count = propertycount) %>%
  dplyr::mutate(date = lubridate::dmy(date)) %>%
  dplyr::rename(latitude = lattitude,
                longitude = longtitude) %>%
  # let's create monthly quarters
  mutate(yr_qtr = as.ordered(lubridate::quarter(date, 
                                             with_year = TRUE,
                                             fiscal_start = 1))) %>%
  # drop price
  tidyr::drop_na(price) %>%
  # make price the log of price
    mutate(price = log10(price))

@@

```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm). There is not much variation amongst the different datasets from the different imputation methods.", fig.width = 9, fig.height = 3, out.width = "100%"}

@@

aug_fits %>%
  ggplot(aes(x = .fitted,
             y = .resid)) + 
  geom_point(alpha = 0.1) + 
  facet_wrap(~model)

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
  # For running headers, if needed
  short:     "\\pkg{naniar}: Explore missing data"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [kworkflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

housing_raw <-
  readr::read_csv(here::here("data",
                             "melbourne_housing_raw.csv")) %>%
  janitor::clean_names() %>%
  dplyr::rename(region_name = regionname,
         property_count = propertycount) %>%
  dplyr::mutate(date = lubridate::dmy(date)) %>%
  dplyr::rename(latitude = lattitude,
                longitude = longtitude) %>%
  # let's create monthly quarters
  mutate(yr_qtr = as.ordered(lubridate::quarter(date, 
                                             with_year = TRUE,
                                             fiscal_start = 1))) %>%
  # drop price
  tidyr::drop_na(price) %>%
  # make price the log of price
    mutate(price = log10(price))

@@

```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm). These are plotted with hex bins, which are coloured according to the number of points in a given hexagon - the brighter the colour the more points. This reveals that the data sets imputed with knn and linear models have more points clustered around zero, whereas complete cases is more flat.", fig.width = 9, fig.height = 3, out.width = "100%"}

@@

aug_fits %>%
  ggplot(aes(x = .fitted,
             y = .resid)) + 
  # geom_point(alpha = 0.1) + 
  geom_hex() +
  labs(fill = "Count") +
  scale_fill_viridis_c() +
  facet_wrap(~model)
